Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-12 manuscript/main.tex
outer product almost derived

For predictor residualization, we fit a hybrid-on-scalar regression model where the predictor is the PLS score for each observation. Specifically, we fit $\widetilde{W}^{[l]}_i = \hat{\rho}^{[l]}_i\, \delta^{[l]} + E_i^{[l]}$, and then take the residuals as the predictor for the next iteration.
Here, $\delta^{[l]}$ and $E_i^{[l]}$ are the regression coefficient and error terms, respectively.
Performing this pointwisely, as in Section \ref{sec: Naive PLS Algorithm}, is highly inefficient and potentially infeasible for dense or irregular functional data. Moreover, the results may lack smoothness, introducing instability that propagates to subsequent results and compromises the entire algorithm.
Therefore, we employ a basis expansion approach to fit the entire hybrid object in one step. We assume both $\delta^{[l]}$ and $E_i^{[l]}$ are elements of $\tilde{\mathcal{H}}$, as defined in \eqref{def:truncated_hilbert_space}. To ensure the stability of the entire PLS estimation procedure, the estimated regression coefficient $\hat{\delta}^{[l]} = (\hat{\delta}^{[l]}_1, \ldots, \hat{\delta}^{[l]}_K, \hat{\boldsymbol{\chi}}^{[l]} ) \in \tilde{\mathcal{H}}$ should have smooth functional parts, consistent with Section \ref{sec: Regularization of PLS Components}. Consequently, $\delta^{[l]}$ is obtained by minimizing a penalized least squares criterion, where the functional part is penalized by the integrated squared second derivative defined in \eqref{def:roughness_penalty}:
\begin{equation}\label{def:argmin_pensse}
	\hat{\delta}^{[l]} := \arg \min_{\delta \in \tilde{\mathcal{H}}} \operatorname{PENSSE}( \delta),
\end{equation}
where
\begin{equation}\label{def:pensse}
	\operatorname{PENSSE}( \delta)
	:=
	\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
	\,
	\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
	\rangle_{\mathcal{H}}
	+
	\tau
	\sum_{j=1}^K\operatorname{PEN}(\delta_j).
\end{equation}
where $\tau>0$ is a  smoothing parameter controlling the trade-off between fit accuracy and functional regression coefficient smoothness.
The following proposition shows that the solution of the problem \ref{def:argmin_pensse} can be obtained by solving linear equations.

\begin{lemma}[Closed-form  solution]\label{proposition:closed_form_orthgonalization}
	Let us denote
	$
	\hat{\boldsymbol{\rho}}^{[l]} := (\hat{\rho}^{[l]}_1, \ldots, \hat{\rho}^{[l]}_n)^\top.
	%\quad\text{and} \quad		\bar{\rho}^{[l]} := (\hat{\boldsymbol{\rho}}^{[l]})^\top \hat{\boldsymbol{\rho}}^{[l]}.
	$
	The solution to the optimization problem \eqref{def:argmin_pensse}, denoted
	$\hat{\delta}^{[l]}   \in \tilde{\mathcal{H}}$, has basis coefficients
	\begin{equation*}
		\hat{\boldsymbol{\pi}}_j^{[l]}
		=
		\bigl(
		\| 	\hat{\boldsymbol{\rho}}^{[l]}\|_2^2
		B
		+
		\tau
		B^{\prime \prime}
		\bigr)^{-1}
		B
		(\Theta_j^{[l]})^\top
		\hat{\boldsymbol{\rho}}^{[l]}
		,
		~
		j = 1, \ldots, K,
	\end{equation*}
	  and scalar parts given by:
	 \begin{equation*}
	 	\hat{\boldsymbol{\chi} }^{[l]}
	 	=
	 	(\mathbf{Z}^{[l]})^\top  \hat{\boldsymbol{\rho}}^{[l]}.
	 \end{equation*}
\end{lemma}
Proof of Lemma \ref{proposition:closed_form_orthgonalization} is provided in Appendix \ref{section:proof:proposition:closed_form_orthgonalization}.

The $(l+1)$-th iteration's residualized predictor is then computed as:
\begin{equation}\label{algorithm_step:residualization_predictor}
	\widetilde{W}_i^{[l+1]} 
	= 
	\widetilde{W}_i^{[l]}  - \widehat{\rho}_{il}  \widehat{ {\delta}}_{l}.
	%%%%%%%%%%%
\end{equation}
Since $\widetilde{W}_i^{[l]} $ are assumed to have a sample mean of zero, these residuals $\widetilde{W}_i^{[l+1]}$   also have a sample mean of zero.

@@

\section{Proof of Lemma \ref{proposition:closed_form_orthgonalization} }\label{section:proof:proposition:closed_form_orthgonalization}
For notational simplicity, we omit the iteration superscript $[l]$.
 Recall from  \eqref{def:plsscore} that
$\hat{\rho}_i
=
\langle \widetilde{W}_i, \, \hat{\xi} \rangle_{\mathcal{H}}$.
Using the basis expansion coefficient notation from \eqref{def:truncated_zeta_and_delta}, \eqref{def:coef_vector_form}, \eqref{def:theta}, and \eqref{def:B_matrix}, the full vector of PLS scores $\hat{\boldsymbol{\rho}}  = (\hat{\rho}_1, \ldots, \hat{\rho}_n)^\top$ 
is computed through the following matrix multiplication:
\begin{equation}\label{score_inner_product_all_n}
	\hat{\boldsymbol{\rho}}^\top = \sum_{k=1}^K ( \hat{\boldsymbol{\gamma}}_j )^\top B \, \Theta_j^\top + \boldsymbol{\zeta}^\top \mathbf{Z}^\top.
\end{equation}
The criterion $\operatorname{PENSSE}( \delta)$ can be decomposed as follows:

\begin{align*}
	\operatorname{PENSSE}( \delta) &= \sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta, \, \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta \rangle_{\mathcal{H}} + \tau \sum_{j=1}^K\operatorname{PEN}(\delta_j) \\
	&= 
	\underbrace{
		\sum_{i=1}^n \langle \widetilde{W}_i^{[l]}, \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}}_{A} 
	\underbrace{
		-2 \sum_{i=1}^n \hat{\rho}_i^{[l]} \langle \delta, \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}} 
	}_{B_1(\boldsymbol{\delta})}
	\underbrace{
		+ 
		(\hat{\boldsymbol{\rho}}^\top \hat{\boldsymbol{\rho}})
		\langle \delta, \, \delta \rangle_{\mathcal{H}}
	}_{B_2(\boldsymbol{\delta})}
	\underbrace{
		+ 
		\tau \sum_{j=1}^K\operatorname{PEN}(\delta_j)
	}_{B_3(\boldsymbol{\delta})}
\end{align*}
Here, part $A$ does not contain $\delta$, so it does not contribute to the minimization problem.

@@

	2\boldsymbol{\chi} ,

@@

	2\hat{\boldsymbol{\chi} }

@@

	=\mathbf{Z}^\top  \hat{\boldsymbol{\rho}}.

@@

Our next task is to show that 
\begin{equation*}
\Theta^{[l_1] }
\mathbb{B}
\hat{\boldsymbol{\xi}}_{l_2} = \mathbf{0} \in \mathbb{R}^n.
\end{equation*}
Without loss of generality, we show that the $i$th entry of $\Theta^{[l_1] }
\mathbb{B}
\hat{\boldsymbol{\xi}}_{l_2}$ is 0.
From the construction in \eqref{def:full_theta}, \eqref{def:theta} and \eqref{def:gram_block}, the $i$-th entry is
\begin{align*}
	(\Theta^{[l_1] }
	\mathbb{B}
	\hat{\boldsymbol{\xi}}_{l_2})_i
	&=
	(\theta_{i11}^{[l_1]}, \ldots, \theta_{i1M}^{[l_1]}, \ldots, \theta_{iK1}^{[l_1]}, \ldots, \theta_{iKM}^{[l_1]}, Z_{i1}^{[l_1]}, \ldots, Z_{ip}^{[l_1]})
	\mathbb{B}
	\hat{\boldsymbol{\xi}}_{l_2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\\&=
	\sum_{k=1}^K
	\boldsymbol{\theta}_{ik}^{[l_1]} B
	\hat{\boldsymbol{\gamma}}_k^{[l_2]}
	+
	\mathbf{Z}_{i}^{[l_1] \top } \hat{\boldsymbol{\zeta}}^{[l_2]}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\	&=
\langle
\widetilde{W}_i^{[l_1]}, 
\hat{\xi}^{[l_2]}
\rangle_{\mathcal{H}},
\end{align*}
where the last equality uses the computation of the $i$th entry in \eqref{score_inner_product_all_n}.

We can express the $l_1$-th residualized hybrid predictor in terms of its previous $(l_1-1)$-th version:
\begin{equation*}
	\widetilde{W}_i^{[l_1]}= \widetilde{W}_i^{[l_1-1]} - \widehat{{\rho}}_i^{[l_1-1]} \widehat{\delta}^{[l_1-1]}
\end{equation*}
\end{proof}

@@

	Let us denote
$
\hat{\boldsymbol{\rho}}^{[l]} := (\hat{\rho}^{[l]}_1, \ldots, \hat{\rho}^{[l]}_n)^\top.
%\quad\text{and} \quad		\bar{\rho}^{[l]} := (\hat{\boldsymbol{\rho}}^{[l]})^\top \hat{\boldsymbol{\rho}}^{[l]}.
$
The solution to the optimization problem \eqref{def:argmin_pensse}, denoted
$\hat{\delta}^{[l]}   \in \tilde{\mathcal{H}}$, has basis coefficients
\begin{equation*}
	\hat{\boldsymbol{\pi}}_j^{[l]}
	=
	\bigl(
	\| 	\hat{\boldsymbol{\rho}}^{[l]}\|_2^2
	B
	+
	\tau
	B^{\prime \prime}
	\bigr)^{-1}
	B
	(\Theta_j^{[l]})^\top
	\hat{\boldsymbol{\rho}}^{[l]}
	,
	~
	j = 1, \ldots, K,
\end{equation*}
and scalar parts given by:
\begin{equation*}
	\hat{\boldsymbol{\chi} }^{[l]}
	=
	(\mathbf{Z}^{[l]})^\top  \hat{\boldsymbol{\rho}}^{[l]}.
\end{equation*}

>>>>

For predictor residualization, we fit a hybrid-on-scalar regression model where the predictor is the PLS score for each observation. Specifically, we fit $\widetilde{W}^{[l]}_i = \hat{\rho}^{[l]}_i\, \delta^{[l]} + E_i^{[l]}$, and then take the residuals as the predictor for the next iteration.
Here, $\delta^{[l]}$ and $E_i^{[l]}$ are the regression coefficient and error terms, respectively.
Performing this pointwisely, as in Section \ref{sec: Naive PLS Algorithm}, is highly inefficient and potentially infeasible for dense or irregular functional data. Moreover, the results may lack smoothness, introducing instability that propagates to subsequent results and compromises the entire algorithm.
Therefore, we employ a basis expansion approach to fit the entire hybrid object in one step. We assume both $\delta^{[l]}$ and $E_i^{[l]}$ are elements of $\tilde{\mathcal{H}}$, as defined in \eqref{def:truncated_hilbert_space}. %To ensure the stability of the entire PLS estimation procedure, the estimated regression coefficient $\hat{\delta}^{[l]} = (\hat{\delta}^{[l]}_1, \ldots, \hat{\delta}^{[l]}_K, \hat{\boldsymbol{\chi}}^{[l]} ) \in \tilde{\mathcal{H}}$ should have smooth functional parts, consistent with Section \ref{sec: Regularization of PLS Components}.
 Consequently, $\delta^{[l]}$ is obtained by minimizing a %penalized 
 least squares criterion:%, where the functional part is penalized by the integrated squared second derivative defined in \eqref{def:roughness_penalty}:
 \begin{equation}\label{def:argmin_pensse}
 	\hat{\delta}^{[l]} := \arg \min_{\delta \in \tilde{\mathcal{H}}} \sum_{i=1}^n 
 	\| \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta\|_{\mathcal{H}}^2.
 \end{equation}
 \begin{comment}
\begin{equation}\label{def:argmin_pensse}
	\hat{\delta}^{[l]} := \arg \min_{\delta \in \tilde{\mathcal{H}}} \operatorname{PENSSE}( \delta),
\end{equation}
where
\begin{equation}\label{def:pensse}
	\operatorname{PENSSE}( \delta)
	:=
	\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
	\,
	\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
	\rangle_{\mathcal{H}}
	+
	\tau
	\sum_{j=1}^K\operatorname{PEN}(\delta_j).
\end{equation}
where $\tau>0$ is a  smoothing parameter controlling the trade-off between fit accuracy and functional regression coefficient smoothness.
\end{comment}
The following proposition shows that the solution of the problem \ref{def:argmin_pensse} can be obtained by solving linear equations.

\begin{lemma}[Closed-form  solution]\label{proposition:closed_form_orthgonalization}
	Let us denote
	$
	\hat{\boldsymbol{\rho}}^{[l]} := (\hat{\rho}^{[l]}_1, \ldots, \hat{\rho}^{[l]}_n)^\top.
	%\quad\text{and} \quad		\bar{\rho}^{[l]} := (\hat{\boldsymbol{\rho}}^{[l]})^\top \hat{\boldsymbol{\rho}}^{[l]}.
	$
	The solution to the optimization problem \eqref{def:argmin_pensse}, denoted
	$\hat{\delta}^{[l]}   \in \tilde{\mathcal{H}}$, has basis coefficients
	\begin{equation*}
		\hat{\boldsymbol{\pi}}_j^{[l]}
		=
		\frac{1}{\| 	\hat{\boldsymbol{\rho}}^{[l]}\|_2^2}
		\Theta_j^{[l] \top}
		\hat{\boldsymbol{\rho}}^{[l]}
		,
		~
		j = 1, \ldots, K,
	\end{equation*}
	  and scalar parts given by:
	 \begin{equation*}
	 	\hat{\boldsymbol{\chi} }^{[l]}
	 	=
	\frac{1}{\| 	\hat{\boldsymbol{\rho}}^{[l]}\|_2^2}
		\mathbf{Z}^{[l] \top}  \hat{\boldsymbol{\rho}}^{[l]}.
	 \end{equation*}
\end{lemma}
Proof of Lemma \ref{proposition:closed_form_orthgonalization} is provided in Appendix \ref{section:proof:proposition:closed_form_orthgonalization}.
\jongmin{proof needs to be revised}
The $(l+1)$-th iteration's residualized predictor is then computed as:
\begin{equation}\label{algorithm_step:residualization_predictor}
	\widetilde{W}_i^{[l+1]} 
	= 
	\widetilde{W}_i^{[l]}  - \widehat{\rho}_{il}  \widehat{ {\delta}}_{l}.
	%%%%%%%%%%%
\end{equation}
Since $\widetilde{W}_i^{[l]} $ are assumed to have a sample mean of zero, these residuals $\widetilde{W}_i^{[l+1]}$   also have a sample mean of zero.

@@

\section{Proof of Lemma \ref{proposition:closed_form_orthgonalization} }\label{section:proof:proposition:closed_form_orthgonalization}
For notational simplicity, we omit the iteration superscript $[l]$.
 Recall from  \eqref{def:plsscore} that
$\hat{\rho}_i
=
\langle \widetilde{W}_i, \, \hat{\xi} \rangle_{\mathcal{H}}$.
Using the basis expansion coefficient notation from \eqref{def:truncated_zeta_and_delta}, \eqref{def:coef_vector_form}, \eqref{def:theta}, and \eqref{def:B_matrix}, the full vector of PLS scores $\hat{\boldsymbol{\rho}}  = (\hat{\rho}_1, \ldots, \hat{\rho}_n)^\top$ 
is computed through the following matrix multiplication:
\begin{equation}\label{score_inner_product_all_n}
	\hat{\boldsymbol{\rho}}^\top = \sum_{k=1}^K ( \hat{\boldsymbol{\gamma}}_j )^\top B \, \Theta_j^\top + \boldsymbol{\zeta}^\top \mathbf{Z}^\top.
\end{equation}
The criterion $\operatorname{PENSSE}( \delta)$ can be decomposed as follows:
\begin{equation}\label{def:pensse}
	\operatorname{PENSSE}( \delta)
	:=
	\sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta,
	\,
	\widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta
	\rangle_{\mathcal{H}}
	+
	\tau
	\sum_{j=1}^K\operatorname{PEN}(\delta_j).
\end{equation}
\begin{align*}
	\operatorname{PENSSE}( \delta) &= \sum_{i=1}^n \langle \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta, \, \widetilde{W}_i^{[l]} - \hat{\rho}_i^{[l]}\, \delta \rangle_{\mathcal{H}} + \tau \sum_{j=1}^K\operatorname{PEN}(\delta_j) \\
	&= 
	\underbrace{
		\sum_{i=1}^n \langle \widetilde{W}_i^{[l]}, \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}}}_{A} 
	\underbrace{
		-2 \sum_{i=1}^n \hat{\rho}_i^{[l]} \langle \delta, \, \widetilde{W}_i^{[l]} \rangle_{\mathcal{H}} 
	}_{B_1(\boldsymbol{\delta})}
	\underbrace{
		+ 
		(\hat{\boldsymbol{\rho}}^\top \hat{\boldsymbol{\rho}})
		\langle \delta, \, \delta \rangle_{\mathcal{H}}
	}_{B_2(\boldsymbol{\delta})}
	\underbrace{
		+ 
		\tau \sum_{j=1}^K\operatorname{PEN}(\delta_j)
	}_{B_3(\boldsymbol{\delta})}
\end{align*}
Here, part $A$ does not contain $\delta$, so it does not contribute to the minimization problem.

@@

	2(\hat{\boldsymbol{\rho}}^\top \hat{\boldsymbol{\rho}})\boldsymbol{\chi} ,

@@

	2 (\hat{\boldsymbol{\rho}}^\top \hat{\boldsymbol{\rho}}) \hat{\boldsymbol{\chi} }

@@

	\frac{1}{(\hat{\boldsymbol{\rho}}^\top \hat{\boldsymbol{\rho}})}
	\mathbf{Z}^\top  \hat{\boldsymbol{\rho}}.

@@

Our next task is to show that 
\begin{equation*}
\Theta^{[l_1] }
\mathbb{B}
\hat{\boldsymbol{\xi}}_{l_2} = \mathbf{0} \in \mathbb{R}^n.
\end{equation*}
From the construction in \eqref{def:full_theta}, \eqref{def:theta} and \eqref{def:gram_block}, the
 $i$th entry of $\Theta^{[l_1] }
\mathbb{B}
\hat{\boldsymbol{\xi}}_{l_2}$ 
is
\begin{align*}
	(\Theta^{[l_1] }
	\mathbb{B}
	\hat{\boldsymbol{\xi}}_{l_2})_i
	&=
	(\theta_{i11}^{[l_1]}, \ldots, \theta_{i1M}^{[l_1]}, \ldots, \theta_{iK1}^{[l_1]}, \ldots, \theta_{iKM}^{[l_1]}, Z_{i1}^{[l_1]}, \ldots, Z_{ip}^{[l_1]})
	\mathbb{B}
	\hat{\boldsymbol{\xi}}_{l_2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\\&=
	\sum_{k=1}^K
	\boldsymbol{\theta}_{ik}^{[l_1]} B
	\hat{\boldsymbol{\gamma}}_k^{[l_2]}
	+
	\mathbf{Z}_{i}^{[l_1] \top } \hat{\boldsymbol{\zeta}}^{[l_2]}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\	&\overset{(i)}{=}
\langle
\widetilde{W}_i^{[l_1]}, 
\hat{\xi}^{[l_2]}
\rangle_{\mathcal{H}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\	&\overset{(ii)}{=}
\langle
\widetilde{W}_i^{[l_1-1]}  , 
\hat{\xi}^{[l_2]}
\rangle_{\mathcal{H}}
-
\widehat{{\rho}}_i^{[l_1-1]}
\langle
   \widehat{\delta}^{[l_1-1]}, 
\hat{\xi}^{[l_2]}
\rangle_{\mathcal{H}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
\\&\overset{(iii)}{=}
\langle
\widetilde{W}_i^{[l_1-1]}  , 
\hat{\xi}^{[l_2]}
\rangle_{\mathcal{H}}
-
\frac{\widehat{{\rho}}_i^{[l_1-1]}}{\| \widehat{\boldsymbol{\rho}}^{[l_1-1]} \|_2^2}
\widehat{\boldsymbol{\rho}}^{[l_1-1] \top}
(
\sum_{k=1}^K {\Theta}_{k}^{[l_1-1]} 
B \boldsymbol{\gamma}_k^{[l_2]}
+
Z^{[l_1-1]} 
\boldsymbol{\zeta}^{[l_2]}
)
%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{align*}
where 
step $(i)$ uses the computation of the $i$th entry in \eqref{score_inner_product_all_n},
step $(ii)$ uses $
	\widetilde{W}_i^{[l_1]}= \widetilde{W}_i^{[l_1-1]} - \widehat{{\rho}}_i^{[l_1-1]} \widehat{\delta}^{[l_1-1]}
$
 and step $(iii)$ uses 
 the closed-form formula of $\widehat{\delta}^{[l_1-1]}$ given by Lemma \ref{proposition:closed_form_orthgonalization}.
	Therefore, we have
\begin{equation*}
\Theta^{[l_1] }
\mathbb{B}
\hat{\boldsymbol{\xi}}_{l_2}
=
\langle
\widetilde{W}^{[l_1-1]}  , 
\hat{\xi}^{[l_2]}
\rangle_{\mathcal{H}}
-
\frac{1}{\| \widehat{\boldsymbol{\rho}}^{[l_1-1]} \|_2^2}
\widehat{\boldsymbol{\rho}}^{[l_1-1] \top}
\langle
\widetilde{W}^{[l_1-1]}  , 
\hat{\xi}^{[l_2]}
\rangle_{\mathcal{H}}
\widehat{\boldsymbol{\rho}}^{[l_1-1]}
\end{equation*}
	\begin{equation*}
	\hat{\boldsymbol{\pi}}_j^{[l]}
	=
	\frac{1}{\| 	\hat{\boldsymbol{\rho}}^{[l]}\|_2^2}
	\Theta_j^{[l] \top}
	\hat{\boldsymbol{\rho}}^{[l]}
	,
	~
	j = 1, \ldots, K,
\end{equation*}
and scalar parts given by:
\begin{equation*}
	\hat{\boldsymbol{\chi} }^{[l]}
	=
	(\mathbf{Z}^{[l]})^\top  \hat{\boldsymbol{\rho}}^{[l]}.
\end{equation*}

@@
